Cognitive Principles and Individual Differences \\ in Human Syllogistic Reasoning

### 2019-11-21 paper/paper-v3.tex
final

Even though the main logical formalism of Classical Logic (CL) has successfully been applied to our scientific reasoning, Kotarbi{\'n}ski~\cite{kotarbinski:1913} already stated in 1913, \textit{the need to revise the two-valued logic as it seemed to interfere with the freedom of human thinking}~\cite[p.17]{malinowski:1993}.
% Originally, one of the objectives in the field of AI and logic programming was to formalize human reasoning and commonsense reasoning~\cite{McCarthy59,McCarthy98}. 
% However, the proposed approaches have been strongly critized~(cf.~\cite{mcdermott:87}).
% On the other hand, cognitive scientists have extensively investigated and acquired a great expertise
% in human thinking. Recently, several attempts have been made to bridge the gap among these disciplines. 
% % as they were mostly theoretical and did not apply to real case studies.
% However, only recently an attempt was made 
% this expertise was not consulted, i.e. there was no communication among the relevant fields. 
% but providing a unified, integrated  accepted theory still appears 
% to be a major challenge. 
In the last decades, cognitive scientists have repeatedly
shown through experimental studies that human reasoning systematically deviates from conclusions that are valid under CL (cf.~\cite{wason:68,byrne:89}),
and thus they seem to agree with Kotarbi{\'n}ski: Classical (two-valued) logic is not an adequate candidate for a cognitive theory.
 What should be the properties of such a candidate in the first place?

@@

As the above example and recent investigations on psychological experiments show (cf.~\cite{khemlani:2016, ragni:2017}), it seems that different humans represent and reason about conditionals differently.
This implies that \textit{the} human reasoner does not exist, instead there 
seem to exist various groups of reasoners, i.e.\ reasoning clusters. The identification of these differences and the specification of these clusters is central 
for the development of an adequate cognitive theory. Only recently approaches have been proposed that account for individual differences (or clusters) in syllogistic reasoning (cf.~\cite{khemlani:2016,declare:2017}).

@@

Here, we provide a qualitative assessment of the \textit{clustering by principles} approach proposed in~\cite{declare:2017},
that performed well on aggregated data, and investigate whether these clusters are also adequate with respect to the non-aggregated dataset given by Johnson-Laird and Khemlani~\cite{khemlani:2016}.
In Section~\ref{sect:wcsclustering} and~\ref{sect:representing}, we introduce the underlying computational logic approach, the cognitive principles and their formalization. Section~\ref{sect:aggregateddata} illustrates an example and briefly discusses the performance of the approach presented in~\cite{declare:2017} on non-aggregated data.
Section~\ref{sect:nonag} provides a more adequate characterization of the individuals' reasoning patterns.

@@

\begin{table}[t]
\[\small
\begin{array}[c]{@{\hspace{0mm}}c|c}
F & \neg F\\ \midrule
\top & \bot \\
\bot & \top \\
\udf & \udf
\end{array}
\;\
\begin{array}{c|lll}
 \wedge & \top & \udf & \bot \\
\midrule
\top & \top & \udf & \bot \\
\udf & \udf & \udf & \bot \\
\bot & \bot & \bot & \bot
 \end{array}
\;\
\begin{array}{c|lll}
 \vee & \top & \udf & \bot \\
\midrule
\top & \top & \top & \top \\
\udf & \top & \udf & \udf \\
\bot & \top & \udf & \bot
 \end{array}
\;\
\begin{array}{c|lll}
\leftarrow & \top & \udf & \bot \\
\midrule
\top & \top & \top & \top \\
\udf & \udf & \top & \top \\
\bot & \bot & \udf & \top
 \end{array}
\;\
\begin{array}{c|ccc}
  \leftrightarrow  & \top & \udf & \bot \\
  \midrule
\top & \top & \udf & \bot \\
\udf & \udf & \top & \udf\\
\bot & \bot & \udf & \top
 \end{array}
\;\
\begin{array}{c|c}
 L & \cxt(L)\footnotemark\\
\midrule
\top & \top \\
\bot & \bot  \\
\udf & \bot
\end{array}
\]
\caption{The truth tables for the connectives under the three-valued {\L}ukasiewicz logic and for $\cxt$. $L$ is a literal and $\top$, $\bot$, and $\udf$ denote \textit{true}, \textit{false},
and \textit{unknown}, respectively.\label{tab:3vldluka}}
\end{table}
\footnotetext{Thanks to the
 anonymous reviewer for the observation, that the $\cxt$ operator seems to be a special case of {\L}ukasiewicz necessity operator~\cite[p. 20]{malinowski:1993}, restricted to literals.
 }

@@

\subsection{On aggregated Data}\label{sect:aggregateddata}

@@

\subsection{On non-aggregated Data} \label{sect:nonag}

@@

\subsubsection{Individual Reasoning Patterns}

@@

Taking the non-aggregated data from~\cite{khemlani:2016} as benchmark,
we investigated how well their responses could be predicted through the cognitive principles,
and whether the individual participants' responses would consistently correspond to a certain set of principles. Recall, their dataset consisted of responses from 20 participants, with each two entries,
as the task was carried out twice, with a break of one week.

@@

We identified three clusters, using the same labels as in~\cite{khemlani:2016}: 
The last two columns show the classification of the participants as deliberative, intuitive or intermediate. 
% The participants in the deliberative cluster here corresponds exactly to~\cite{khemlani:2016}. 
Our classification only differs with respect to Johnson-Laird and Khemlani's 
classification~\cite{khemlani:2016} for the intermediate and intuitive cluster with respect to three participants (1, 11 and 19).
A possible explanation for this difference could be that they
did not take into account the participants' changes between week 1 and week 2.

Two of the three clusters can be specified through a set of cognitive principles, but for one
group of participants, no distinguishable pattern could be identified:
% The summary is shown in Table~\ref{tab:clusters}.
\begin{description}
 \item[Deliberative cluster] (6 out of 20 participants) is characterized by the participants whose responses corresponded to the predictions made by basic/ contraposition. Almost none of their conclusions corresponded to predictions made by abduction/ generalization. Theses participants did change their responses from week 1 to week 2 
 lower than average and concluded higher than average NVC. Generally, their responses improved from week 1 to week 2 with respect to CL.
 \item[Intuitive cluster] (7 out of 20 participants)
is characterized by the participants whose responses corresponded to the predictions given by abduction/ generalization. Almost none of their conclusions corresponded to the predictions made by contraposition (except for 2 participants in week 2). They showed a high change within their own responses from week 1 to week 2 
 and concluded lower than average NVC. Generally, their responses improved from week 1 to week~2 with respect to CL, but they never performed \textit{as well as} the participants in 
 the deliberative cluster.
 \item[Intermediate cluster] (7 out of 20 participants) is characterized by the participants whose responses did not corresponded to any particular set of principles. These participants had a high change within their own responses from week 1 to week~2
 and no particular pattern with respect to NVC conclusions. Generally they were not improving their responses from week 1 to week 2 with respect to CL, but normally \textit{compensated} for responses valid in CL in some cases (that corresponded to contraposition) with responses not valid in CL in the other cases (that correspond to abduction/ generalization).
\end{description}

>>>>

Even though the main logical formalism of Classical Logic (CL) has successfully been applied to our scientific reasoning, Kotarbi{\'n}ski~\cite{kotarbinski:1913} already stated in 1913 \textit{the need to revise the two-valued logic as it seemed to interfere with the freedom of human thinking}~\cite[p.17]{malinowski:1993}.
% Originally, one of the objectives in the field of AI and logic programming was to formalize human reasoning and commonsense reasoning~\cite{McCarthy59,McCarthy98}. 
% However, the proposed approaches have been strongly critized~(cf.~\cite{mcdermott:87}).
% On the other hand, cognitive scientists have extensively investigated and acquired a great expertise
% in human thinking. Recently, several attempts have been made to bridge the gap among these disciplines. 
% % as they were mostly theoretical and did not apply to real case studies.
% However, only recently an attempt was made 
% this expertise was not consulted, i.e. there was no communication among the relevant fields. 
% but providing a unified, integrated  accepted theory still appears 
% to be a major challenge. 
In the last decades, cognitive scientists have repeatedly
shown through experimental studies that human reasoning systematically deviates from conclusions that are valid under CL (cf.~\cite{wason:68,byrne:89}),
and thus they seem to agree with Kotarbi{\'n}ski: Classical (two-valued) logic is not an adequate candidate for a cognitive theory.
 But what should be the properties of such a candidate in the first place?

@@

As the above example and recent investigations on psychological experiments show (cf.~\cite{khemlani:2016, ragni:2017}), it seems that different humans represent and reason about conditionals differently.
This implies that \textit{the} human reasoner does not exist, instead there 
seem to exist various groups of reasoners, i.e.\ reasoning clusters. The identification of these differences and the specification of these clusters is central 
for the development of an adequate cognitive theory. Only recently approaches have been proposed that account for individual differences in syllogistic reasoning (cf.~\cite{khemlani:2016,declare:2017}).

@@

Here, we provide a qualitative assessment of the \textit{clustering by principles} approach proposed in~\cite{declare:2017},
that performed well on aggregated data, and investigate whether these clusters are also adequate with respect to the non-aggregated dataset provided by Johnson-Laird and Khemlani~\cite{khemlani:2016}.
In Section~\ref{sect:wcsclustering} and~\ref{sect:representing}, we introduce the underlying computational logic approach, the cognitive principles and their formalization. Section~\ref{sect:aggregateddata} illustrates an example and briefly discusses the performance of the approach presented in~\cite{declare:2017} on non-aggregated data.
Section~\ref{sect:nonag} provides novel characterization of the individuals' reasoning patterns through cognitive principles.

@@

\begin{table}[t]
\[\small
\begin{array}[c]{@{\hspace{0mm}}c|c}
F & \neg F\\ \midrule
\top & \bot \\
\bot & \top \\
\udf & \udf
\end{array}
\;\
\begin{array}{c|lll}
 \wedge & \top & \udf & \bot \\
\midrule
\top & \top & \udf & \bot \\
\udf & \udf & \udf & \bot \\
\bot & \bot & \bot & \bot
 \end{array}
\;\
\begin{array}{c|lll}
 \vee & \top & \udf & \bot \\
\midrule
\top & \top & \top & \top \\
\udf & \top & \udf & \udf \\
\bot & \top & \udf & \bot
 \end{array}
\;\
\begin{array}{c|lll}
\leftarrow & \top & \udf & \bot \\
\midrule
\top & \top & \top & \top \\
\udf & \udf & \top & \top \\
\bot & \bot & \udf & \top
 \end{array}
\;\
\begin{array}{c|ccc}
  \leftrightarrow  & \top & \udf & \bot \\
  \midrule
\top & \top & \udf & \bot \\
\udf & \udf & \top & \udf\\
\bot & \bot & \udf & \top
 \end{array}
\;\
\begin{array}{c|c}
 L & \cxt(L)\footnotemark\\
\midrule
\top & \top \\
\bot & \bot  \\
\udf & \bot
\end{array}
\]
\caption{The truth tables for the connectives under the three-valued {\L}ukasiewicz logic and for $\cxt$. $L$ is a literal and $\top$, $\bot$, and $\udf$ denote \textit{true}, \textit{false},
and \textit{unknown}, respectively.\label{tab:3vldluka}}
\end{table}
\footnotetext{Thanks to the
 anonymous reviewer for the observation, that the $\cxt$ operator seems to be a special case of {\L}ukasiewicz necessity operator (cf.~\cite[p. 20]{malinowski:1993}), restricted to literals.
 }

@@

\subsection{Clusters}\label{sect:aggregateddata}

@@

\subsection{Individual Reasoners} \label{sect:nonag}

@@

\subsubsection{Characterization through Principles}

@@

Taking the non-aggregated data from~\cite{khemlani:2016},
we investigated how well the participants' responses could be explained through the cognitive principles,
and whether their individual responses would be consistent with respect to a certain set of principles. Recall, Johnson-Laird and Khemlani's dataset consisted of responses from 20 participants, with each two entries,
as the task was carried out twice, with a break of one week.

@@

We identified three types of reasoners, characterizing them according to the terms already used in~\cite{khemlani:2016}: 
The last two columns show the classification of the participants as deliberative, intuitive or intermediate. 
% The participants in the deliberative cluster here corresponds exactly to~\cite{khemlani:2016}. 
Our classification only differs with respect to Johnson-Laird and Khemlani's 
classification~\cite{khemlani:2016} for the intermediate and intuitive reasoners with respect to three participants (1, 11 and 19).
A possible explanation for this difference could be that they
did not take into account the participants' changes between week 1 and week 2.
Two of the three types of reasoners can be characterized through a set of cognitive principles, but for one
group of participants, no distinguishable pattern could be identified:
% The summary is shown in Table~\ref{tab:clusters}.
\begin{description}
 \item[Deliberative reasoners] (6 out of 20 participants) is characterized by the participants whose responses corresponded to the predictions made by basic/ contraposition. Almost none of their conclusions corresponded to predictions made by abduction/ generalization. Theses participants did change their responses from week 1 to week 2 
 lower than average and concluded higher than average NVC. Generally, their responses improved from week 1 to week 2 with respect to CL.
 \item[Intuitive reasoners] (7 out of 20 participants)
is characterized by the participants whose responses corresponded to the predictions given by abduction/ generalization. Almost none of their conclusions corresponded to the predictions made by contraposition (except for 2 participants in week 2). They showed a high change within their own responses from week 1 to week 2 
 and concluded lower than average NVC. Generally, their responses improved from week 1 to week~2 with respect to CL, but they never performed \textit{as well as} the participants in 
 the deliberative cluster.
 \item[Intermediate reasoners] (7 out of 20 participants) is characterized by the participants whose responses did not corresponded to any particular set of principles. These participants had a high change within their own responses from week 1 to week~2
 and no particular pattern with respect to NVC conclusions. Generally they were not improving their responses from week 1 to week 2 with respect to CL, but normally \textit{compensated} for responses valid in CL in some cases (that corresponded to contraposition) with responses not valid in CL in the other cases (that correspond to abduction/ generalization).
\end{description}
